Analysis of finite-volume advection schemes on cubed-sphere grids and an accurate \textcolor{red}{scheme} for divergent winds

### 2024-05-20 paper.tex
modified distance notation, grid imprinting explanantion

The transport scheme employed in FV3 extends the method introduced by \cite{lin:1996} to the cubed-sphere, as proposed by \cite{putman:2007}. 
This approach involves constructing a two-dimensional (2D) scheme by combining the solution of one-dimensional (1D) conservative transport equations using a direction-splitting strategy on each cube face.
The 1D equations are solved using the finite-volume approach of the Piecewise Parabolic Method (PPM) \cite{colella:1984,carpenter:1990}. 
A notable feature of the scheme proposed by \cite{lin:1996} is its elimination of the splitting error under conditions where the initial transported scalar field density is constant and the wind is divergence-free. 
This property is attained through modifications to the inner advection operators employed within the scheme.

The goal of this study is to reassess and suggest enhancements for the current FV3 transport scheme originally developed by \cite{putman:2007}, given its pivotal role in the horizontal dynamics of FV3. 
We demonstrate that the scheme proposed by \cite{putman:2007} assumes constant metric terms during the application of the PPM to each 1D flux integration domain and employs a first-order departure point calculation for the 1D fluxes.
We propose the development of a scheme that incorporates a second-order departure point calculation for the 1D fluxes and eliminates the assumption of constant metric terms. 
While the proposed scheme does not retain the property of splitting error elimination for divergence-free winds and constant scalar fields, it introduces only second-order errors in such scenarios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equi-edge mapping}
\label{cs-equiedge}
Another cubed-sphere mapping is the equi-edge proposed by
\cite{chen:2021} using $\beta(x) = \sqrt{2}\tan{x}$ and
$a=\arcsin{\big(\frac{1}{\sqrt{3}}\big)}$.
The idea behind the equi-edge mapping lies in partitioning the edges of the spherical cube equally,
and then generating the other cells, hence the name equi-edge.
Also, this mapping leads to more uniform grid cells after applying the grid stretching option of FV3 \citep{harris:2016, chen:2021}.
\textcolor{red}{This mapping is illustrated in Figure \ref{sph-cube-equiedge}}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textcolor{red}{
The following geodesic distances in $x$ and $y$ directions, respectively, are introduced:
\begin{align}
	\label{distcube}
	\widehat{d^x_{ij}} = d(\Psi_{i+\frac{1}{2},j},\Psi_{i-\frac{1}{2},j}) , \quad
	\widehat{d^y_{ij}} = d(\Psi_{i,j+\frac{1}{2}},\Psi_{i,j-\frac{1}{2}}),
\end{align}
where $d(P,Q) = R\arccos{(\langle P, Q \rangle)}$, for $P,Q \in \mathbb{S}^2_R$.}
These distances may be represented in terms of the tangent vector norms as:
\begin{align}
	\widehat{d^x_{ij}} = 
	\int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}}
	\|\partial_x  {\Psi}_{p}\|(x,y_j) \,dx ,\quad
	\widehat{d^y_{ij}} =
	\int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}
	\|\partial_y {\Psi}_{p}\|(x_i,y) \,dy.
\end{align}
Hence, their midpoint approximations are defined as:
\begin{align}
	\label{distcube2}
	{d^x_{ij}} = \|\partial_x {\Psi}_{p}(x_i,y_j)  \|\Delta x,\quad
	{d^y_{ij}} = \|\partial_y {\Psi}_{p}(x_i,y_j) \|\Delta y.
\end{align}
We point out that ${d^x_{ij}}$ and ${d^y_{ij}}$ are replaced in FV3 code by the geodesic distances that they approximate whenever they appear, which are second-order accurate by the midpoint rule.
A control volume of the cubed-sphere is denoted by $\Omega_{ijp}$, defined as $\Omega_{ijp} = {\Psi}_p(\Omega_{ij})$.
The area of $\Omega_{ijp}$ is denoted by $|\Omega_{ij}|$,
since the area does not depend on the panel due to the grid symmetry.
The control volume area may be expressed as:
\begin{equation}
	\label{area}
	|\Omega_{ij}| = \int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}} \int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}{\sqrt{\mathfrak{g}}(x,y)} \,dx \,dy = 
	|\hat{\Omega}_{ij}| + \mathcal{O}(\Delta x^2),
\end{equation}
where 
\begin{equation}
	\label{area2}
	|\hat{\Omega}_{ij}| = \sqrt{\mathfrak{g}}_{ij} \Delta x \Delta y,
\end{equation} $\sqrt{\mathfrak{g}}_{ij}=\sqrt{\mathfrak{g}}(x_i,y_j) $
and the last equality in Equation \eqref{area} follows from the midpoint rule for integration.
Similar to the grid lengths, the approximated areas $|\hat{\Omega}_{ij}|$ are replaced by the exact area $|\Omega_{ij}|$ in the FV3 code.
\begin{table}[htbp]
	\centering
	\caption{Mean length, minimum length, and maximum length for different values of $N$ considering the equiangular grid.\label{g2-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline
		48 & 220 & 202 & 240 & 1.1890 \\
		96 & 109 & 99 & 118  & 1.1892 \\
		192 & 54 & 49 & 59  & 1.1892 \\
		384 & 27 & 24 & 29  & 1.1892 \\
		768 & 13 & 12 & 14  & 1.1892\\
		\hline
	\end{tabular} 
\end{table}
\begin{table}[htbp]
	\centering
	\caption{As Table \ref{g2-dx-table} but considering the equi-edge grid. \label{g0-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline 
		48 & 218 & 175 & 266 & 1.5192 \\
		96 & 108 & 86 & 131 & 1.5195 \\
		192 & 54 & 43 & 65 & 1.5196 \\
		384 & 26 & 21 & 32 & 1.5197 \\
		768 & 13 & 10 & 16 & 1.5197 \\
		\hline
	\end{tabular}
\end{table}

Tables  \ref{g2-dx-table} and \ref{g0-dx-table} display the lengths of the  equiangular and equi-edge grids for $N=48\times 2^k$, 
where $k = 0,\ldots, 4$. These values of $N$ are considered in this work.
It can be observed that in terms of length of the cells, the equi-edge grid is less uniform than the equiangular grid.
Despite this, the equi-edge grid is the operational grid in some applications of FV3 \cite{harris:2021,chen:2021}, such as, for instance, 
the Next Generation Global Prediction System (NGGPS) \cite{zhou:2019}, 
because this grid is expected to produce less grid imprinting due to its greater uniformity near the cubed edges.

@@

\subsection{The time-averaged and upwind CFL number}
\label{timeav-cfl}
In order to compute the integral in Equation \eqref{numerical-flux}, it will be useful to introduce the time-averaged CFL number at the edges.
This integral in FV3 is expressed in terms of this number.
For the LT2 scheme, the CFL number at the edges is defined naturally as
\begin{align}
	\label{cfl_LT2}
	\tilde{c}_{i+\frac{1}{2}}^{LT2} = {\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}{\Delta x}
	= {\tilde{u}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}
	{{d^x_{i+\frac{1}{2}}}},
\end{align}
where ${\tilde{u}}_{i+\frac{1}{2}}^{LT2}$ is just a normalization of 
${\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}$ (Equation \eqref{contra-uv}).
 FV3, on the other hand, uses the upwind CFL approach from \cite{lin:1994}, namely:
\begin{align}
	\label{cfl_fv3}
	{c}_{i+\frac{1}{2}}^{FV3}
    ={\tilde{u}}_{i+\frac{1}{2}}^{FV3}\frac{{\Delta t}}
	{{{d^{x,*}_{i+\frac{1}{2}}}}},
\end{align}
where
\begin{align}
	\label{cfl_dp1-dx}
    {{d^{x,*}_{i+\frac{1}{2}}}} = 
	\begin{cases}
		{{d^{x}_{i}}},
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} \ge 0},\\
		{{d^{x}_{i+1}}}, 
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} < 0}.
	\end{cases}
\end{align}
Then Equation \eqref{dp-estimated} may is expressed in terms of the CFL number as:
\begin{equation}
	\label{dp-estimated-cfl}
	\tilde{x}_{i+\frac{1}{2}}^n = x_{i+\frac{1}{2}} - {\tilde{c}}^{M}_{i+\frac{1}{2}}\Delta x,
\end{equation}
where $M=\text{FV3}$ or $M=\text{LT2}$. 
It is easy to see that assuming the absolute value of the CFL number is less than one, then $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i+\frac{1}{2}},{x}_{i+\frac{3}{2}}]$ if ${\tilde{c}}^{M}_{i+\frac{1}{2}}$ is positive, and $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i-\frac{1}{2}},{x}_{i+\frac{1}{2}}]$ otherwise.
Then, as mentioned earlier before, the integral of Equation \eqref{numerical-flux} is constrained to a single control-volume under the CFL condition.
Also, the upwind approach of the numerical flux becomes clear at this point.

@@

\bigg( \frac{{d}^x_{i+\frac{1}{2}}}{{d}^{x,*}_{i+\frac{1}{2}}}\bigg)

@@

		d^y_{i+\frac{1}{2}} 

@@

and recalling the definition of ${{d}^{x,*}_{i+\frac{1}{2}}}$ in Equation \eqref{cfl_dp1-dx}.

>>>>

The transport scheme employed in FV3 extends the method introduced by \cite{lin:1996} to the cubed-sphere, as proposed by \cite{putman:2007}. 
\textcolor{red}{
	The major challenge of transitioning from latitude-longitude grids to grids based on Platonic solids \cite{stan:2012} is grid imprinting, which refers to the appearance of grid features on the solution (cf. e.g., \cite{peixoto:2013, weller:2012, mouallem:2023}), and is highly undesirable due to its lack of physical meaning.}

This approach involves constructing a two-dimensional (2D) scheme by combining the solution of one-dimensional (1D) conservative transport equations using a direction-splitting strategy on each cube face.
The 1D equations are solved using the finite-volume approach of the Piecewise Parabolic Method (PPM) \cite{colella:1984,carpenter:1990}. 
A notable feature of the scheme proposed by \cite{lin:1996} is its elimination of the splitting error under conditions where the initial transported scalar field density is constant and the wind is divergence-free. 
This property is attained through modifications to the inner advection operators employed within the scheme. 
The goal of this study is to reassess and suggest enhancements for the current FV3 transport scheme originally developed by \cite{putman:2007}, given its pivotal role in the horizontal dynamics of FV3. 
We demonstrate that the scheme proposed by \cite{putman:2007} assumes constant metric terms during the application of the PPM to each 1D flux integration domain and employs a first-order departure point calculation for the 1D fluxes.
We propose the development of a scheme that incorporates a second-order departure point calculation for the 1D fluxes and eliminates the assumption of constant metric terms. 
While the proposed scheme does not retain the property of splitting error elimination for divergence-free winds and constant scalar fields, it introduces only second-order errors in such scenarios.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Equi-edge mapping}
\label{cs-equiedge}
\textcolor{red}{
Another cubed-sphere mapping is the equi-edge mapping, initially introduced by \cite{chen:2021}, which utilizes $\beta(x) = \sqrt{2}\tan{x}$ and $a=\arcsin{\big(\frac{1}{\sqrt{3}}\big)}$. It is worth noting that while this mapping technique had been used previously in FV3, it was not formally documented until the work \cite{chen:2021}.}
The idea behind the equi-edge mapping lies in partitioning the edges of the spherical cube equally,
and then generating the other cells, hence the name equi-edge.
Also, this mapping leads to more uniform grid cells after applying the grid stretching option of FV3 \citep{harris:2016, chen:2021}.
\textcolor{red}{This mapping is illustrated in Figure \ref{sph-cube-equiedge}}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\textcolor{red}{
The following geodesic distances in $x$ and $y$ directions, respectively, are introduced:
\begin{align}
	\label{distcube}
	\widehat{l^x_{ij}} = d(\Psi_{i+\frac{1}{2},j},\Psi_{i-\frac{1}{2},j}) , \quad
	\widehat{l^y_{ij}} = d(\Psi_{i,j+\frac{1}{2}},\Psi_{i,j-\frac{1}{2}}),
\end{align}
where $d(P,Q) = R\arccos{(\langle P, Q \rangle)}$, for $P,Q \in \mathbb{S}^2_R$.}
These distances may be represented in terms of the tangent vector norms as:
\begin{align}
	\widehat{l^x_{ij}} = 
	\int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}}
	\|\partial_x  {\Psi}_{p}\|(x,y_j) \,dx ,\quad
	\widehat{l^y_{ij}} =
	\int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}
	\|\partial_y {\Psi}_{p}\|(x_i,y) \,dy.
\end{align}
Hence, their midpoint approximations are defined as:
\begin{align}
	\label{distcube2}
	{l^x_{ij}} = \|\partial_x {\Psi}_{p}(x_i,y_j)  \|\Delta x,\quad
	{l^y_{ij}} = \|\partial_y {\Psi}_{p}(x_i,y_j) \|\Delta y.
\end{align}
We point out that ${l^x_{ij}}$ and ${l^y_{ij}}$ are replaced in FV3 code by the geodesic distances that they approximate whenever they appear, which are second-order accurate by the midpoint rule.
A control volume of the cubed-sphere is denoted by $\Omega_{ijp}$, defined as $\Omega_{ijp} = {\Psi}_p(\Omega_{ij})$.
The area of $\Omega_{ijp}$ is denoted by $|\Omega_{ij}|$,
since the area does not depend on the panel due to the grid symmetry.
The control volume area may be expressed as:
\begin{equation}
	\label{area}
	|\Omega_{ij}| = \int_{x_{i-\frac{1}{2}}}^{x_{i+\frac{1}{2}}} \int_{y_{j-\frac{1}{2}}}^{y_{j+\frac{1}{2}}}{\sqrt{\mathfrak{g}}(x,y)} \,dx \,dy = 
	|\hat{\Omega}_{ij}| + \mathcal{O}(\Delta x^2),
\end{equation}
where 
\begin{equation}
	\label{area2}
	|\hat{\Omega}_{ij}| = \sqrt{\mathfrak{g}}_{ij} \Delta x \Delta y,
\end{equation} $\sqrt{\mathfrak{g}}_{ij}=\sqrt{\mathfrak{g}}(x_i,y_j) $
and the last equality in Equation \eqref{area} follows from the midpoint rule for integration.
Similar to the grid lengths, the approximated areas $|\hat{\Omega}_{ij}|$ are replaced by the exact area $|\Omega_{ij}|$ in the FV3 code.
\begin{table}[htbp]
	\centering
	\caption{Mean length, minimum length, and maximum length for different values of $N$ considering the equiangular grid.\label{g2-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline
		48 & 220 & 202 & 240 & 1.1890 \\
		96 & 109 & 99 & 118  & 1.1892 \\
		192 & 54 & 49 & 59  & 1.1892 \\
		384 & 27 & 24 & 29  & 1.1892 \\
		768 & 13 & 12 & 14  & 1.1892\\
		\hline
	\end{tabular} 
\end{table}
\begin{table}[htbp]
	\centering
	\caption{As Table \ref{g2-dx-table} but considering the equi-edge grid. \label{g0-dx-table}}
	\begin{tabular}{|c|c|c|c|c|c}\hline
		$N$ & Mean Length (km) & Min Length (km) & Max Length (km) & $\frac{\text{Max}}{\text{Min}}$ \\ \hline 
		48 & 218 & 175 & 266 & 1.5192 \\
		96 & 108 & 86 & 131 & 1.5195 \\
		192 & 54 & 43 & 65 & 1.5196 \\
		384 & 26 & 21 & 32 & 1.5197 \\
		768 & 13 & 10 & 16 & 1.5197 \\
		\hline
	\end{tabular}
\end{table}

Tables  \ref{g2-dx-table} and \ref{g0-dx-table} display the lengths of the  equiangular and equi-edge grids for $N=48\times 2^k$, 
where $k = 0,\ldots, 4$. These values of $N$ are considered in this work.
It can be observed that in terms of length of the cells, the equi-edge grid is less uniform than the equiangular grid.
Despite this, the equi-edge grid is the operational grid in some applications of FV3 \cite{harris:2021,chen:2021}, such as, for instance, 
the Next Generation Global Prediction System (NGGPS) \cite{zhou:2019}, 
because this grid is expected to produce less grid imprinting due to its greater uniformity near the original cube.

@@

\subsection{The time-averaged and upwind CFL number}
\label{timeav-cfl}
In order to compute the integral in Equation \eqref{numerical-flux}, it will be useful to introduce the time-averaged CFL number at the edges.
This integral in FV3 is expressed in terms of this number.
For the LT2 scheme, the CFL number at the edges is defined naturally as
\begin{align}
	\label{cfl_LT2}
	\tilde{c}_{i+\frac{1}{2}}^{LT2} = {\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}{\Delta x}
	= {\tilde{u}}_{i+\frac{1}{2}}^{LT2}\frac{\Delta t}
	{{l^x_{i+\frac{1}{2}}}},
\end{align}
where ${\tilde{u}}_{i+\frac{1}{2}}^{LT2}$ is just a normalization of 
${\mathfrak{\tilde{u}}}_{i+\frac{1}{2}}^{LT2}$ (Equation \eqref{contra-uv}).
 FV3, on the other hand, uses the upwind CFL approach from \cite{lin:1994}, namely:
\begin{align}
	\label{cfl_fv3}
	{c}_{i+\frac{1}{2}}^{FV3}
    ={\tilde{u}}_{i+\frac{1}{2}}^{FV3}\frac{{\Delta t}}
	{{{l^{x,*}_{i+\frac{1}{2}}}}},
\end{align}
where
\begin{align}
	\label{cfl_dp1-dx}
    {{l^{x,*}_{i+\frac{1}{2}}}} = 
	\begin{cases}
		{{l^{x}_{i}}},
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} \ge 0},\\
		{{l^{x}_{i+1}}}, 
		\quad &\text{if }{{{u}}_{i+\frac{1}{2}}^{FV3} < 0}.
	\end{cases}
\end{align}
Then Equation \eqref{dp-estimated} may is expressed in terms of the CFL number as:
\begin{equation}
	\label{dp-estimated-cfl}
	\tilde{x}_{i+\frac{1}{2}}^n = x_{i+\frac{1}{2}} - {\tilde{c}}^{M}_{i+\frac{1}{2}}\Delta x,
\end{equation}
where $M=\text{FV3}$ or $M=\text{LT2}$. 
It is easy to see that assuming the absolute value of the CFL number is less than one, then $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i+\frac{1}{2}},{x}_{i+\frac{3}{2}}]$ if ${\tilde{c}}^{M}_{i+\frac{1}{2}}$ is positive, and $\tilde{x}_{i+\frac{1}{2}}^n \in [{x}_{i-\frac{1}{2}},{x}_{i+\frac{1}{2}}]$ otherwise.
Then, as mentioned earlier before, the integral of Equation \eqref{numerical-flux} is constrained to a single control-volume under the CFL condition.
Also, the upwind approach of the numerical flux becomes clear at this point.

@@

\bigg( \frac{{l}^x_{i+\frac{1}{2}}}{{l}^{x,*}_{i+\frac{1}{2}}}\bigg)

@@

		l^y_{i+\frac{1}{2}} 

@@

and recalling the definition of ${{l}^{x,*}_{i+\frac{1}{2}}}$ in Equation \eqref{cfl_dp1-dx}.
